This practical aims to guide you through some of the key ideas in
**ggplot2**. As with the first practical, feel free to experiment. Some of the functions introduced in this practical haven't been explicitly covered in the notes. Use the built-in R help or the **ggplot2** help pages at

> http://had.co.nz/ggplot2/

\noindent as needed.


## Over plotting

Scatter plots are very useful. However, when we have a large data set, points
will be plotted on top of each other obscuring the relationship. We call this
problem over plotting. There are a few techniques we can use to help, although
the best solution is often problem specific.

To begin with we will create an example data frame:

```{r,overplotting, cache=TRUE, tidy=FALSE, echo=1:4}
## If your computer is slow when plotting reduce the value of n
library("jrGgplot2")
library("ggplot2")
df = overplot_data(n = 20000)
h = ggplot(df) + geom_point(aes(x, y))
```

```{r,echo=FALSE, fig.margin = TRUE, fig.cap="A scatter plot that suffers from over plotting.", fig.height = 4}
h
```

\noindent We can create a simple scatter plot of this data using the following command

```{r, echo = TRUE}
h = ggplot(df) + geom_point(aes(x, y))
```

\noindent This plot isn't particularly good. Try to improve it by using a
combination of:


- changing the transparency level: `alpha`;^[`alpha` takes a value between $0$ and $1$.]
- change the shape: `shape=1` and `shape=`.'`
- use some jittering - `geom_jitter`.
- adding a contour to the plot using `stat_density2d`.
- What does

```{r,cache=TRUE, fig.keep='none', tidy=FALSE, echo = TRUE}
h +  stat_density2d(aes(x, y, fill = ..density..),
                contour = FALSE, geom = "tile")
```
do?    
- What does `stat_bin2d()` and `stat_binhex()` do - add it to the
  plot to find out! Try varying the parameters `bins` and `binwidth`.

Now, let's try to make the Jumping Rivers logo appear from within the large array of scattered points\sidenote{Magic and mysterious, I know}. Load in the data and experiment with the size and alpha aesthetics to display the image. The source code to load the data and produce the ggplot object are given below.

```{R, echo=TRUE, eval=FALSE}
data("xydata", package = "jrGgplot2")
```

```{r, echo=FALSE, warning=FALSE}
library("png")
library("ggplot2")
library("tibble")
library("dplyr")
logo = readPNG("../inst/logo.png")
dims = dim(logo)
pts = matrix(c(rep(1:dims[1], dims[2]), rep(1:dims[2], each = dims[1])), byrow = FALSE, ncol = 2)

#pick out only the points that are grey (on outline)
pts = pts[sapply(1:(dims[1] * dims[2]),
                 function(x) {
                   logo[pts[x, 1], pts[x, 2], 4] > 0.5 &
                   logo[pts[x, 1], pts[x, 2], 4] < 0.8
                 }
                 ), ]

#nolint start
samp = sample(x = 1:nrow(pts), size = 5000, replace = TRUE)
#nolint end

ptsflipped = cbind(x = pts[samp, 2], y = dims[2] - pts[samp, 1])

#convert from matrix to tibble
ptsflipped = as.tibble(ptsflipped)

#centre image
ptsflipped = ptsflipped %>% 
  mutate(x  = x + 300, y = y + 300)

# now add a random scatter of points
xynorm = round(cbind(x = rnorm(n = 20000, mean = 1000, sd = 700),
                     y = rnorm(n = 20000, mean = 1400, sd = 700)))

xydata = rbind(xynorm, ptsflipped)
n = nrow(xydata)

#reorder rows to hide the data
samp2 = sample(x = 1:n, size = n, replace = FALSE)
xydata = xydata[samp2, ]

xydata = as.tibble(xydata)
```

```{r, echo=TRUE, eval=FALSE}
ggplot() +
  geom_point(data = xydata, mapping = aes(x = x, y = y))
```

## Displaying distributions

The diamonds data set contains the prices and other attributes of almost 54,000 diamonds. It is a data frame with $53,940$ rows and $10$ variables. First, load the diamonds data set:

```{r, echo = TRUE}
data(diamonds, package = "ggplot2")
```

\noindent and look at the help file:

```{r,tidy=FALSE, eval=FALSE, echo = TRUE}
?diamonds
```

\noindent We can construct a histogram of diamond depth using the following commands:

```{r,fig.keep='none', message=FALSE, echo = TRUE}
i1 = ggplot(data = diamonds) +
  geom_histogram(aes(x = depth))
```

```{r,echo=FALSE, message=FALSE, fig.margin = TRUE, fig.cap = "Histogram of the diamond data set.", fig.height = 4}
i1
```


\noindent to get figure 2. Let's experiment a bit.

1. Change the `binwidth` in the `geom_histogram`. What value do you
  think is best?
2. What happens when you set `colour=cut` in the `geom_histogram`
  aesthetic? What other options can you change?^[Look at the `geom_histogram` help page: http://had.co.nz/ggplot2/geom_histogram.html}
3. Replace `geom_histogram` with `geom_density`. Set `fill=cut` and change the `alpha` value.
4. Try `geom_boxplot`.



## Copy cat

```{r,F1, echo=FALSE, message=FALSE, fig.keep='none',results='hide'}
data(mpg, package = "ggplot2")
g = ggplot(data = mpg, aes(x = displ, y = hwy))
g1 = g + geom_point() + stat_smooth(linetype = 2) +
  labs(x = "Displacement", y = "Highway mpg")
g2 = g + geom_point() + stat_smooth(aes(colour = drv))
```

```{r,dev='png', out.width='\\textwidth', echo=FALSE, message=FALSE, fig.margin = TRUE, fig.cap = "Graphics for section 1.", fig.height = 4}
g1
```

The aim of this section is to recreate the graphics in figures 3 and 4. Feel
free to experiment. To begin, load the package

```{r,message=FALSE, echo = TRUE}
library("ggplot2")
```

\noindent and the `mpg` data set

```{r,results='hide', echo = TRUE}
data(mpg, package = "ggplot2")
dim(mpg)
```

1. Figure 3: Create a scatter plot of engine displacement,
  `displ`, against highway mpg, `hwy`. To get started:

```{r,fig.keep='none', tidy=FALSE, echo = TRUE}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  labs(x = "Displacement")
```

Now add a dashed loess line and change the $y$-axis label.
Hint: try `stat_smooth` and adding a $y$ label to the previous labs function.

```{r}
g = ggplot(data = mpg, aes(x = displ, y = hwy))
g1 = g + geom_point() + stat_smooth(linetype = 2) +
  labs(x = "Displacement",
       y = "Highway mpg")
```

```{r,dev='png', out.width='\\textwidth', echo=FALSE, message=FALSE, fig.margin = TRUE, fig.cap = "Graphics for section 1.", fig.height = 4}
g2
```
2. Figure 4: Using `stat_smooth`, add a loess line conditional
  on the drive.
```{r}
g2 = g + geom_point() + stat_smooth(aes(colour = drv))
```

## Solutions

Solutions are contained within this package:
```{r,eval=FALSE}
library(jrGgplot2)
vignette("solutions2", package = "jrGgplot2")
```
