This practical aims to guide you through some of the key ideas in
**ggplot2**. As with the first practical, feel free to experiment. Some of the functions introduced in this practical haven't been explicitly covered in the notes. Use the built-in R help or the **ggplot2** help pages at

> http://had.co.nz/ggplot2/

\noindent as needed.


## Over plotting

Scatter plots are very useful. However, when we have a large data set, points
will be plotted on top of each other obscuring the relationship. We call this
problem over plotting. There are a few techniques we can use to help, although
the best solution is often problem specific.

To begin with we will create an example data frame:

```{r,overplotting, cache=TRUE, tidy=FALSE, echo=1:4}
## If your computer is slow when plotting reduce the value of n
library("jrGgplot2")
library("ggplot2")
df = overplot_data(n = 20000)
h = ggplot(df) + geom_point(aes(x, y))
```

```{r,echo=FALSE, fig.margin = TRUE, fig.cap="A scatter plot that suffers from over plotting.", fig.height = 4}
h
```

\noindent We can create a simple scatter plot of this data using the following command

```{r, echo = 2}
library("ggplot2")
h = ggplot(df) + geom_point(aes(x, y))
```

\noindent This plot isn't particularly good. Try to improve it by using a
combination of:


- changing the transparency level: `alpha`;^[`alpha` takes a value between $0$ and $1$.]
- change the shape: `shape=1` and `shape=`.'`
- use some jittering - `geom_jitter`.
- adding a contour to the plot using `stat_density2d`.
- What does

```{r,cache=TRUE, fig.keep='none', tidy=FALSE, echo = TRUE}
h +  stat_density2d(aes(x, y, fill = ..density..),
                contour = FALSE, geom = "tile")
```
do?    
- What does `stat_bin2d()` and `stat_binhex()` do - add it to the
  plot to find out! Try varying the parameters `bins` and `binwidth`.

## Displaying distributions

The diamonds data set contains the prices and other attributes of almost 54,000 diamonds. It is a data frame with $53,940$ rows and $10$ variables. First, load the diamonds data set:

```{r, echo = TRUE}
data(diamonds, package = "ggplot2")
```

\noindent and look at the help file:

```{r,tidy=FALSE, eval=FALSE, echo = TRUE}
?diamonds
```

\noindent We can construct a histogram of diamond depth using the following commands:

```{r,fig.keep='none', message=FALSE, echo = TRUE}
i1 = ggplot(data = diamonds) +
  geom_histogram(aes(x = depth))
```

```{r,echo=FALSE, message=FALSE, fig.margin = TRUE, fig.cap = "Histogram of the diamond data set.", fig.height = 4}
i1
```


\noindent to get figure 2. Let's experiment a bit.

1. Change the `binwidth` in the `geom_histogram`. What value do you
  think is best?
2. What happens when you set `colour=cut` in the `geom_histogram`
  aesthetic? What other options can you change?^[Look at the `geom_histogram` help page: http://had.co.nz/ggplot2/geom_histogram.html}
3. Try `geom_density`. Set `fill=cut` and change the `alpha` value.
4. Try `geom_boxplot`.



## Copy cat


```{r,F1, echo=FALSE, message=FALSE, fig.keep='none',results='hide'}
data(mpg, package = "ggplot2")
g = ggplot(data = mpg, aes(x = displ, y = hwy))
g1 = g + geom_point() + stat_smooth(linetype = 2) +
  xlab("Displacement") + ylab("Highway mpg")
g2 = g + geom_point() + stat_smooth(aes(colour = drv))
```

```{r,dev='png', out.width='\\textwidth', echo=FALSE, message=FALSE, fig.margin = TRUE, fig.cap = "Graphics for section 1.", fig.height = 4}
g1
```

The aim of this section is to recreate the graphics in figures 3 and 4. Feel
free to experiment. To begin, load the package

```{r,message=FALSE, echo = TRUE}
library("ggplot2")
```

\noindent and the `mpg` data set

```{r,results='hide', echo = TRUE}
data(mpg, package = "ggplot2")
dim(mpg)
```

1. Figure 3: Create a scatter plot of engine displacement,
  `displ`, against highway mpg, `hwy`. To get started:

```{r,fig.keep='none', tidy=FALSE, echo = TRUE}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point() + xlab("Displacement")
```

Now add a dashed loess line and change the $y$-axis label.
Hint: try `stat_smooth` and `ylab(`New label')`.

```{r}
g = ggplot(data = mpg, aes(x = displ, y = hwy))
g1 = g + geom_point() + stat_smooth(linetype = 2) +
  xlab("Displacement") + ylab("Highway mpg")
```

```{r,dev='png', out.width='\\textwidth', echo=FALSE, message=FALSE, fig.margin = TRUE, fig.cap = "Graphics for section 1.", fig.height = 4}
g2
```
2. Figure 4: Using `stat_smooth`, add a loess line conditional
  on the drive.
```{r}
g2 = g + geom_point() + stat_smooth(aes(colour = drv))
```

## Solutions

Solutions are contained within this package:
```{r,eval=FALSE}
library(jrGgplot2)
vignette("solutions2", package = "jrGgplot2")
```
